---
title: "Homework, week 12: Population cycling"
author: "Gera"
date: "1/20/2020"
output: html_document
---

In this report we explore a fake dataset consisting of the total number of mice and foxes on a island for 100 years. Here we use a Lotka-Voltera-type model to fit the data in which mice reproduce, but are eaten by foxes, and fox reproduction rate depending on how many mice they eat.  Here we hope to get a fit for a model of how next year’s number mice and foxes by using the current year’s number. The statistical mode here is built by the use of Stan which performs full Bayesian inference. 



```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, echo=FALSE, warning=FALSE, message=FALSE)

library(ggplot2)
library(tidyr)

library("rstan")


dataset <- read.table("/Users/gerardoperez/Documents/shell/Bi610/UO_ABS/CLASS_MATERIALS/Homeworks/mice_and_foxes.tsv", header=T)
head(dataset)

summary(dataset)

#ggplot() + geom_line(aes(x=dataset$year,y=dataset$mice),color='red') + 
           geom_line(aes(x=dataset$year,y=dataset$foxes),color='blue') 
          
year<-dataset$year
mice<-dataset$mice
foxes<-dataset$foxes
plot(year, mice, type="o", col="blue", pch="o",ylab="Total number",main="Figure 1: Mice vs Foxes")
points(year, foxes, col='red')
lines(year, foxes, col="red")
legend("topleft",legend=c("mice","foxes"), col=c("blue","red"),  lty=1:2, cex=0.8)



lot_volt_block <- "
data {
    int N;
    int F[N];
    int M[N];
}
parameters {
    real epsilon;
    real gamma;
    real<lower=0, upper=1> p;
    real<lower=0> lambda;
}
model {
    vector[N-1] r;
    epsilon ~ beta(1,1);
    r=exp(-epsilon*to_vector(F)[1:(N-1)]);
    M[2:N] ~ poisson((lambda + r).*to_vector(M)[1:(N-1)]);
    F[2:N] ~ poisson((p*to_vector(F)[1:(N-1)]+gamma*(1-r)).*to_vector(F)[1:(N-1)]);
}
"
lot_volt_model <- stan_model(model_code=lot_volt_block)
osc_fit <- sampling(lot_volt_model,
                    data=list(N=nrow(dataset),
                              F=dataset[,2],
                              M=dataset[,3]),
                    iter=1000, chains=3,
                    control=list(max_treedepth=12))



#stan_hist(osc_fit)

#plot(osc_fit)

#stan_plot(osc_fit)
#summary(osc_fit)
#plot(osc_fit, show_density = FALSE, ci_level = 0.5, outer_level = 0.95)
#summary(osc_fit)



```

First we explore the data by doing statistics to the data as shown in Table 1.    


**Table 1: Summary statitics on dataset**
```{r}

lot_volt_block <- "
data {
    int N;
    int F[N];
    int M[N];
}
parameters {
    real epsilon;
    real gamma;
    real<lower=0, upper=1> p;
    real<lower=0> lambda;
}
model {
    vector[N-1] r;
    epsilon ~ beta(1,1);
    r=exp(-epsilon*to_vector(F)[1:(N-1)]);
    M[2:N] ~ poisson((lambda + r).*to_vector(M)[1:(N-1)]);
    F[2:N] ~ poisson((p*to_vector(F)[1:(N-1)]+gamma*(1-r)).*to_vector(F)[1:(N-1)]);
}
"
summary(dataset)


```

The years in the  data has a min of 1751 and a max of 1850, which indicates the timeline of 1751 to 1850. Also, the mice have a mean of 618.9 with a min of 109 and a max 1839. Finally, the foxes have a mean of 338.2 with a min of 42 and a max 1082.  

The criteria for a Lotka-Voltera equations are shrinkage of predator population, growth of prey population, large prey population results to a large predator population, and large predator population results to a smaller prey population (Bob Carpenter, 2018). We plot the dataset of prey and predator, as shown in Figure 1, and we can see the Lotka-Voltera criteria which results in mice and foxes oscillating as the years go by.

```{r}
plot(year, mice, type="o", col="blue", pch="o",ylab="Total number",main="Figure 1: Mice vs Foxes")
points(year, foxes, col='red')
lines(year, foxes, col="red")
legend("topleft",legend=c("mice","foxes"), col=c("blue","red"),  lty=1:2, cex=0.8)

```

Since the data consist of counts we will use a Poisson distribution for our model. We used a link exponential link function, called r,  to calculate the chance that a mouse escapes all foxes when there are foxes. We let epsilon be the per fox encounter rate, lambda be the per capita mouse fecundity, p be the probability of survival until the next year for each fox, and gamma be the conversion rate from mice eaten to baby foxes. In addition we added a non-informative Beta(1,1)  prior, which is flat for all possible values of our parameter epsilon. So our stan model ended up looking:


lot_volt_block <- "

data {

    int N;
    
    int F[N];
    
    int M[N];
    
}

parameters {

    real epsilon;
    
    real gamma;
    
    real<lower=0, upper=1> p;
    
    real<lower=0> lambda;
    
}

model {

    vector[N-1] r;
    
    epsilon ~ beta(1,1);
    
    r=exp(-epsilon*to_vector(F)[1:(N-1)]);
    
    M[2:N] ~ poisson((lambda + r).*to_vector(M)[1:(N-1)]);
    
    F[2:N] ~ poisson((p*to_vector(F)[1:(N-1)]+gamma*(1-r)).*to_vector(F)[1:(N-1)]);
    
}
"


Running this model resulted in histograms for posterior analysis. Our parameters epsilon and p did not look great as shown in Figure 2

**Figure 2: Histograms for posterior analysis**
```{r}

stan_hist(osc_fit)

```

```{r}
true_osc <- list(ep = summary(osc_fit)$summary[1,1], l = summary(osc_fit)$summary[2,1], p = summary(osc_fit)$summary[3,1], g = summary(osc_fit)$summary[4,1])
N = nrow(osc_fit)
xy <- matrix(nrow=N, ncol=2)
xy[1,] <- c(250,800)
for (k in 1:(N-1)) {
    xy[k+1,] <- c(rpois(1,(true_osc$l + exp(-true_osc$ep * xy[k,2])) * xy[k,1]) ,
                  rpois(1,(true_osc$p * xy[k,2]) + (true_osc$g * (1 - exp(-true_osc$ep * xy[k,2])) * xy[k,1])))
}
xy <- as.data.frame(xy) 
colnames(xy) <- c("mice", "foxes")
xy$year <- dataset$year


```

We also looked at our uncertainty by looking at the 95% credible intervals for our parameters as shown in Figure 3.

```{r}
plot(osc_fit, show_density = FALSE, ci_level = 0.5, outer_level = 0.95, fill_color = "salmon")+ ggtitle("Figure 3: 95% credible intervals for parameters ")


```

To get exact confidence intervals we used a summary that is shown in Table 2.

**Table 2: Summary of  the fit model**
```{r, message=FALSE, warning=FALSE}


summary(osc_fit)



```

Parameter epsilon had a mean of 
0.5 with a 95% confidence interval of 0.008 to 0.9. parameter gamma had a mean of 1 with a small 95% confidence interval 1 to 1. Parameter p had a mean 3E-7 with a small 95% confidence interval of 1E-8 to 1E-6. Parameter lambda had a mean of 1 with a small 95% confidence interval of 1 to 1. 




We then did a simulation with the dataset under our model using your inferred parameter values and plotted against the dataset. Figure 4 shows mice and Figure 5 shows the foxes. 

```{r, message=FALSE, warning=FALSE}







year<-dataset$year[1:100]

plot(year, dataset$mice[1:100], type="o", col="blue", pch="o",ylab="Total number",main="Figure 4: Mice True vs Mice Model")
points(year, xy$mice[1:100], col='purple')
lines(year, xy$mice[1:100], col="purple")
legend("topleft",legend=c("True","Model"), col=c("blue","purple"),  lty=1:2, cex=0.8)






```

```{r, message=FALSE, warning=FALSE}

plot(year, dataset$foxes[1:100], type="o", col="red", pch="o",ylab="Total number",main="Figure 5: Foxes True vs Foxes Model")
points(year, xy$foxes[1:100], col='brown')
lines(year, xy$foxes[1:100], col="brown")
legend("topleft",legend=c("True","Model"), col=c("red","brown"),  lty=1:2, cex=0.8)

```


The two figures show that our model was not a good model since the simulation and true don’t look similar. This can be a result of a bad set up our parameters, priors or the model. We will further work on improving this model





**Reference**

Carpenter, Bob,. Predator-Prey Population Dynamics: 
the Lotka-Volterra model in Stan, Columbia University in New York. 8 January 2018.)
https://mc-stan.org/users/documentation/case-studies/lotka-volterra-predator-prey.html
